\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-08-26 paper.qmd
Added main idea for Section 6, need to clean the description and add everthing in a nice flow

@@

XXX Add tSNE, UMAP, and PaCMAP NLDR layouts and describe what is happening
XXX Discuss the models of one specific cluster (filled out, pancakes): tSNE, UMAP, PaCMAP are useful to capture global structures, but this is an evidence where PaCMAP fail to capture the local structure with default hyper-parameter settings than tSNE and UMAP
XXX Why? hyper-parameter choice specially number of neighbors, data characteristics (intrinsic dimensionality), imbalance between global and local structures, optimization and initialization issues
XXX Add this is with specific hyper-parameter choices

>>>>

In the t-SNE 2D layout, the distances between clusters appear smaller, leading to more tightly packed clusters, which is typical of t-SNE's focus on preserving local structures. This often results in less emphasis on the global distances between clusters. On the other hand, UMAP tends to maintain larger separations between clusters, better reflecting the global structure and preserving the relative distances between clusters more effectively. PaCMAP, meanwhile, balances these aspects by maintaining the distance between clusters and positioning them correctly, effectively capturing both global and local structures.

The t-SNE, UMAP, and PaCMAP layouts each reveal different aspects of the clustering structure within the dataset. When examining a specific cluster, which appears more like a filled-out pancake in higher dimensions, we observe that t-SNE, UMAP, and PaCMAP all effectively capture global structures. However, there is evidence suggesting that PaCMAP, with its default hyper-parameter settings, struggles to preserve the local structure as effectively as t-SNE and UMAP.

The differences in performance can be attributed to several factors. One key element is the choice of hyper-parameters, particularly the number of neighbors considered during the dimensionality reduction process. This parameter significantly impacts how well local structures are maintained. Additionally, the intrinsic dimensionality of the data, the balance between global and local structures, and issues related to optimization and initialization all contribute to the observed outcomes.

It's important to note that these observations are made with specific hyper-parameter settings. Adjusting these parameters could lead to different results, potentially improving PaCMAP's ability to capture local structures or altering how t-SNE and UMAP handle global and local structures. Thus, while PaCMAP generally balances global and local preservation, careful tuning of its parameters is crucial for achieving the best results in any given dataset.

@@

<!-- XXX Add tSNE, UMAP, and PaCMAP NLDR layouts and describe what is happening -->
<!-- XXX Discuss the models of one specific cluster (filled out, pancakes): tSNE, UMAP, PaCMAP are useful to capture global structures, but this is an evidence where PaCMAP fail to capture the local structure with default hyper-parameter settings than tSNE and UMAP -->
<!-- XXX Why? hyper-parameter choice specially number of neighbors, data characteristics (intrinsic dimensionality), imbalance between global and local structures, optimization and initialization issues -->
<!-- XXX Add this is with specific hyper-parameter choices -->

### 2024-08-26 paper.tex
Added 2D layouts of Five Gau
Added main idea for Section 6, need to clean the description and add everthing in a nice flow

XXX Add tSNE, UMAP, and PaCMAP NLDR layouts and describe what is
happening XXX Discuss the models of one specific cluster (filled out,
pancakes): tSNE, UMAP, PaCMAP are useful to capture global structures,
but this is an evidence where PaCMAP fail to capture the local structure
with default hyper-parameter settings than tSNE and UMAP XXX Why?
hyper-parameter choice specially number of neighbors, data
characteristics (intrinsic dimensionality), imbalance between global and
local structures, optimization and initialization issues XXX Add this is
with specific hyper-parameter choices

>>>>

In the t-SNE 2D layout, the distances between clusters appear smaller,
leading to more tightly packed clusters, which is typical of t-SNE's
focus on preserving local structures. This often results in less
emphasis on the global distances between clusters. On the other hand,
UMAP tends to maintain larger separations between clusters, better
reflecting the global structure and preserving the relative distances
between clusters more effectively. PaCMAP, meanwhile, balances these
aspects by maintaining the distance between clusters and positioning
them correctly, effectively capturing both global and local structures.

The t-SNE, UMAP, and PaCMAP layouts each reveal different aspects of the
clustering structure within the dataset. When examining a specific
cluster, which appears more like a filled-out pancake in higher
dimensions, we observe that t-SNE, UMAP, and PaCMAP all effectively
capture global structures. However, there is evidence suggesting that
PaCMAP, with its default hyper-parameter settings, struggles to preserve
the local structure as effectively as t-SNE and UMAP.

The differences in performance can be attributed to several factors. One
key element is the choice of hyper-parameters, particularly the number
of neighbors considered during the dimensionality reduction process.
This parameter significantly impacts how well local structures are
maintained. Additionally, the intrinsic dimensionality of the data, the
balance between global and local structures, and issues related to
optimization and initialization all contribute to the observed outcomes.

It's important to note that these observations are made with specific
hyper-parameter settings. Adjusting these parameters could lead to
different results, potentially improving PaCMAP's ability to capture
local structures or altering how t-SNE and UMAP handle global and local
structures. Thus, while PaCMAP generally balances global and local
preservation, careful tuning of its parameters is crucial for achieving
the best results in any given dataset.

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-five-gau-nldr-layouts-1.pdf}

\caption{\label{fig-five-gau-nldr-layouts}(a) tSNE, (b) UMAP, and (c)
PaCMAP}
